The onset of of hydrodynamically driven waving of marine grass in flow

### 2014-09-26 Paper/grass.tex
Minor language fix
Incorporate all of Maha's suggestions

\begin{document}
\title{Linear instability underlying synchronous waving of marine grass in flow}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
\author{L. Mahadevan}
\affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi\thanks{work performed while visiting Brown University}}
\affiliation{Collective Interactions Unit, OIST Graduate University, 1919-1 Tancha, Onna-son, Okinawa, Japan 904-0495}
\author{Amala Mahadevan}
\affiliation{Woods Hold Institute of Oceanography, Woods Hole MA USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

The flexibility of the grass blades, on the other hand, serves merely aid in visualizing the flow structure resulting from the instability\cite{Nepf2012}. 

@@

where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, $N_g$ the grass number density per unit area. In the canopy the equations of continuity \& momentum balance can be written as 

@@

 -{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0

@@

A comparison of laboratory mean profile with the steady flow profile is shown in figure \eqref{basicflow}.
Flow within grass has approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from the balance between drag and with pressure gradient. 
Flow outside has a simple parabolic velocity profile due to the balance between viscous forces and the pressure gradient. 

@@

This dependence of the boundary layer thickness on the grass density $N_g$, shown in Figure \ref{basicflow}, gives us a way to systematically control it and investigate its effect on the instability mechanism.

@@

Considering small perturbations $u, v, p$ about the base profile $U$ and $P$ given by the solution of \eqref{base_equ}. Substituting them in equation \eqref{navier-stokes} and expanding  to linear order in the perturbation variable we obtain.

@@

The growth rate $\sigma$ for a given wave number $k=2\pi /\lambda$ can be shown to satisfy the eigenvalue problem.

@@

\begin{table*}
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.4}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{$-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& $\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{1/2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto \delta^{-1}$		& $\propto \delta^{-1}$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& ~~~~yes, near $y=\hg$~~~~			& no
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the mechanism of waving and the origin of the threshold, we consider the behavior of the instability as a function of the vegetation drag.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical grass density discontinuously jumps and remains $O(1)$ (see figure \ref{Re_vs_delta}). 
To aid in explaining the behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in figure \ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two. We refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the grass density increases, causing the discontinuity in the most unstable mode.

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits an asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold. 
This limit can be understood through asymptotically estimating the sizes of the terms in \eqref{Orr-somerfield} using $D\sim \delta^{-1}$, $\sigma \sim \delta^{-1}$, and $U\sim \delta$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
The terms balance when $\Rey \sim {\Ndg}^{1/2}$ (or $\Rey \sim H/\delta$) leading to a simplification of \eqref{Orr-somerfield} as
\begin{align}
(D^2-k^2)^2 \phi- 2 \Rey \Ndg D( SUD\phi) = \Rey \sigma (D^2-k^2)\phi,
\end{align} 
in a region of O($\delta$) near $y=\hg$.
Since the relative magnitude of the terms on the r.h.s. to those on the l.h.s. of this equation is $\Rey/\Ndg^{1/2}$, we expect identical asymptotic behavior for fixed $\Rey/\Ndg^{1/2}$. Therefore the threshold obtained for Mode 1 is $\Rey \propto \Ndg^{1/2}$ explaining the numerically observation (see figure \ref{Re_vs_delta}). 
This analysis also concludes that the mode structure is self-similar over the length scale $\delta$ for fixed $\Rey/\Ndg^{1/2}$; the verification of this expectation is shown in figure \ref{Asymptotic_mode}, supporting this argument.

On the other hand Mode 2 has a threshold $\Rey \sim ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in figure \ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of \eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but fixed $\Rey/\Ndg \sim O(1)$.
Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the grass is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
This estimate simplifies \eqref{Orr-somerfield} to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,  \quad &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, \quad &\text{ for $y>\hg$}.
\end{split}
\end{align}
\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of grass density. 
Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The mode shapes approach each other for these small values of $\delta/H$ indicating that the large grass density asymptote is reached. Mode 1 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. 
Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. 
The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the grass height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}
Since $\Rey/\Ndg$ is the only remaining parameter, the mode shape would converge in the aforementioned limit, in agreement with our numerical results for Mode 2 shown in figure \ref{Asymptotic_mode}. 
We interpret Mode 2 as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the grass tip plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).
%On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.

@@

From these asymptotic behaviors, Mode 1 may be interpreted as the instability of the flow in the boundary layer, whereas Mode 2 may be understood as the instability on the scale of the water column. 

@@

@@

The appearance of vegetation drag parameter in the dominant balance of \eqref{Orr-somerfield} and the threshold criteria demonstrates its role in setting the threshold.

@@

Despite these differences, in the framework of our model, the limit of $\hg/H \ll 1$ while $\delta/\hg$ = constant can be used to represents the hydrodynamic instability for the terrestrial case with stiff grass blades. We find that in this case, the transition from Mode 1 to Mode 2 happens at such a large grass density, so as to make Mode 2 irrelevant. We recover the Kelvin-Helmholtz-like characteristics observed in the terrestrial case. 

>>>>

\begin{document}
\title{The onset of of hydrodynamically driven waving of marine grass in flow}
\author{Ravi Singh}
\affiliation{Brown University, Providence RI 02912 USA}
\author{L. Mahadevan}
\affiliation{Harvard University, Cambridge MA 02138 USA}
\author{M. M. Bandi\thanks{work performed while visiting Brown University}}
\affiliation{Collective Interactions Unit, OIST Graduate University, 1919-1 Tancha, Onna-son, Okinawa, Japan 904-0495}
\author{Amala Mahadevan}
\affiliation{Woods Hold Institute of Oceanography, Woods Hole MA USA}
\author{Shreyas Mandre}
\affiliation{Brown University, Providence RI 02912 USA}

@@

The flexibility of the grass blades, on the other hand, serves merely to aid in visualizing the flow structure resulting from the instability\cite{Nepf2012}. 

@@

where $\mathbf{f_{d}}$ is drag force per unit length of the grass blade, $N_g$ the grass number density per unit area\cite{Vivoni98,Nepf99,Ghisal02,Delangre04,Delangre06}. 
In the canopy the equations of continuity and momentum balance can be written as 

@@

 -\frac{dP}{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0

@@

A comparison of the steady flow profile from the solution of \eqref{base_equ} with experimental measurements is shown in figure~\ref{basicflow}.
Flow within grass has approximately constant velocity $U_g \sim \sqrt{\frac{dP/dx}{\rho C_N dN_g}}$ arising from the balance between drag and with pressure gradient, and the flow outside has a simple parabolic velocity profile due to the balance between viscous forces and the pressure gradient. 

@@

This dependence of the boundary layer thickness on the grass density $N_g$, shown in Figure \ref{basicflow}, gives us a way to systematically investigate its effect on the instability mechanism.

@@

To understand the evolution of small perturbations to the steady flow profile, we substitute $\bu = (U+\tilde{u}, \tilde{v})$, $p=P+\tilde{p}$ in \eqref{navier-stokes} and expand to linear order to obtain (after dropping the tilde decoration)

@@

The growth rate $\sigma$ for a given wave number $k=2\pi /\lambda$ satisfies the eigenvalue problem.

@@

\begin{table*}
\rowcolors{3}{tableShade}{white}  %% start alternating shades from 3rd row
\renewcommand{\arraystretch}{1.4}
 \begin{tabular}{l|c|c|c}
			& Kelvin-Helmholtz 				& Mode 1 		& Mode 2 \\ \hline
 Base velocity profile 	& $U(y) = U_0 \tanh(y/\delta)$			& \multicolumn{2}{c}{$-{dP}/{dx}+\mu U''(y) +S(y) \rho C_N d N_gU^2=0$} \\
 Domain 		& $-\infty < y < \infty$			& \multicolumn{2}{c}{$-1<y<1$} \\
 Inflection point	& exists at $y=0$				& \multicolumn{2}{c}{$U''(y)$ discontinuous at $y=\hg$} \\
 Shear layer thickness	& $\delta$					& \multicolumn{2}{c}{$\delta \sim  H\left(\Rey \Ndg \right)^{-1/3}$} \\
 Linearized dynamics	& $\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi$		& \multicolumn{2}{c}{Equation \eqref{Orr-somerfield}} \\
 Critical parameters	& none						& $\Rey \propto \Ndg^{1/2}$ 	& $\Rey \propto \Ndg$ \\
 Most unstable $k$ as $\delta \to 0$	& $\propto \delta^{-1}$		& $\propto \delta^{-1}$	& $O(1)$ \\
 Mode localized?	& yes, near $y=0$				& ~~~~yes, near $y=\hg$~~~~			& no
 \end{tabular}
 \caption{Comparison between Kelvin-Helmholtz instability and the two modes resulting from solution of \ref{Orr-somerfield}.}
 \label{tab:comparison}
\end{table*}
To better understand the mechanism of waving and the origin of the threshold, we consider the behavior of the instability as a function of the vegetation drag characterized by $\Ndg$.
The fastest growing wavenumber first increases proportional to $H/\delta$, but at a critical grass density discontinuously jumps and remains $O(1)$ (see figure \ref{Re_vs_delta}). 
To aid in explaining the behavior, we show heat maps of Re$(\sigma)$ as a function of $\Rey$ and $k$, for different $\hg/H$ and $\Ndg$ in figure \ref{K_Re_sigma_set3}. 
The lowest $\Rey$ on the neutral curve sets the threshold. 
We observe that as $\Ndg$ increases, the unstable region splits into two. We refer to the unstable region with the higher $k$ as ``Mode 1'', and the one with the lower $k$ as ``Mode 2''. 
The unstable region for Mode 1, depending on $\hg/H$ either recedes to higher $\Rey$ or shrinks to zero size, as the grass density increases, causing the most unstable mode to jump.

The distinct asymptotic behavior of the two modes as $\Ndg \gg 1$ allows us to understand the mechanism of the instability. 
Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits an asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold. 
This limit can be understood through asymptotically estimating the sizes of the terms in \eqref{Orr-somerfield} using $D\sim \delta^{-1}$, $\sigma \sim \delta^{-1}$, and $U\sim \delta$; the magnitude of the advection term is $\Rey/\delta^3 \sim \Rey^2 \Ndg $, and the viscous term (or the vegetation drag term, which is found to scale identically) in the boundary layer is $\delta^{-4} \sim (\Rey \Ndg)^{4/3}$. 
The terms balance when $\Rey \sim {\Ndg}^{1/2}$ (or $\Rey \sim H/\delta$) leading to a simplification of \eqref{Orr-somerfield} as
\begin{align}
(D^2-k^2)^2 \phi- 2 \Rey \Ndg D( SUD\phi) = \Rey \sigma (D^2-k^2)\phi,
\label{eqn:mode1asymp}
\end{align} 
in a region of thickness O($\delta$) near $y=\hg$.
Since the relative magnitude of the terms on the r.h.s. to those on the l.h.s. of this equation is $\Rey/\Ndg^{1/2}$, we expect identical asymptotic behavior for fixed $\Rey/\Ndg^{1/2}$. Therefore the threshold obtained for Mode 1 is $\Rey \propto \Ndg^{1/2}$ (or $\Rey \propto H/\delta$) explaining the numerically observed asymptote (see figure \ref{Re_vs_delta}). 
This analysis also concludes that the mode structure is self-similar over the length scale $\delta$ for fixed $\Rey/\Ndg^{1/2}$; the verification of this expectation is shown in figure \ref{Asymptotic_mode}, supporting this argument.

On the other hand, the threshold condition for Mode 2 is numerically observed to be $\Rey \propto ({\delta}/{H})^{-3/2}$ (or $\Rey \propto \Ndg$) for $k\sim O(1)$, shown in figure \ref{Re_vs_delta}, which can be understood by an asymptotic scaling analysis of \eqref{Orr-somerfield} by assuming $\Rey \gg 1$ but fixed $\Rey/\Ndg \sim O(1)$.
Note that for large $\Ndg$, the non-dimensional steady flow velocity inside the grass is $U_g \sim (\Rey \Ndg)^{-1/2}$. 
This estimate simplifies \eqref{Orr-somerfield} to 
\begin{align}
\begin{split}
\sigma\left(D^2-k^2\right)\phi = -2{(\Ndg/\Rey)^{1/2}}D^2\phi,  \quad &\text{ for $y<\hg$}  \\
\left(\sigma+ikU\right) \left(D^2-k^2\right)\phi =  ikU_{yy}\phi, \quad &\text{ for $y>\hg$}.
\end{split}
\label{eqn:mode2asymp}
\end{align}
\begin{figure}
\includegraphics[]{Asymptotic_noshear}
\caption{Plot of mode shape $|\phi|$ for different shear layer thickness in limit of small shear layer thickness for representative values of grass density. 
Mode 1 is shown in solid and Mode 2 is shown in dashed. The parameters Mode 2 shapes are chosen such that $\Rey \gg 1$, $\Ndg \gg 1$ (specified in terms of $\delta/H$) but $\Rey/\Ndg = O(1)$. 
The mode shapes approach each other for these small values of $\delta/H$ indicating that the large grass density asymptote is reached. Mode 1 shapes appear self-similar in shape as $\delta\to 0$, 
and are compared to each other on a rescaled coordinate in the inset. 
Inset shows $|\phi|$ for Mode 1 as a function of $(y-\hg)/\delta$. The modes approach a universal shape, indicating that an asymptotic limit has been reached. 
The limit is not yet reached for the case $\delta/H = 0.005$ due to the influence of bottom boundary; the grass height in this case is comparable to the boundary layer thickness.}
\label{Asymptotic_mode}
\end{figure}
Since $\Rey/\Ndg$ is the only remaining parameter, the mode shape would converge in the aforementioned limit, in agreement with our numerical results for Mode 2 shown in figure \ref{Asymptotic_mode}. 
We interpret Mode 2 as the instability of an inviscid flow, with the vegetation modeled by a continuum drag field, and for which the boundary layer near the grass tip plays no role. The only remaining parameter $\Rey/\Ndg$ 
sets the threshold, leading to the asymptotic behavior $\Rey \propto \Ndg$ (or $\Rey \sim ({\delta}/{H})^{-3/2}$).
%On the other hand, Mode 1 asymptotically localizes to the boundary layer near the grass tip, and exhibits a different asymptotic behavior with $k \sim O(H/\delta)$, and $\Rey \sim (H/\delta)$ (or $\Rey \propto {\Ndg}^{1/2}$) at the threshold.

@@

@@

Because Mode 1 eigenfunction is localized over a length scale $\delta$, it may be interpreted as the instability of the flow in the boundary layer, whereas Mode 2 may be understood as the instability on the scale of the water column. 

@@

The appearance of vegetation drag parameter in the dominant balances \eqref{eqn:mode1asymp} and \eqref{eqn:mode2asymp}, and the threshold criteria demonstrates its role in setting the threshold.

@@

Despite these differences, in the framework of our model, the limit of $\hg/H \ll 1$ while $\delta/\hg$ = constant can be used to represents the hydrodynamic instability for the terrestrial case with stiff grass blades. We find that in this case, the transition from Mode 1 to Mode 2 happens at such a large grass density, so as to make Mode 2 irrelevant. In this manner, we recover the Kelvin-Helmholtz-like characteristics observed in the terrestrial case. 
